Pangenome graph layout by Path-Guided Stochastic Gradient Descent

### 2023-09-22 manuscript/fig_1d_sorts.tex
each sentence in one line
integrate feedback from Niklas

\begin{figure*}[!htpb]
	\centering
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.png}
		\label{fig:1d_fig1}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.png}
		\label{fig:1d_fig2}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.Y.og.png}
		\label{fig:1d_fig3}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.HY.og.png}
		\label{fig:1d_fig4}
	\end{subfigure}	
	\caption{\textit{odgi viz} 1D visualizations of a 5 haplotypes subgraph of the Human Pangenome Resource Consortium (HPRC) chromosome 6 pangenome graph. The major histocompatibility complex (MHC) sequence was injected as an extra path. Various node arrangements are shown. 
		\textbf{a-d} A graphs nodes are arranged from left to right forming the pangenome sequence. The black lines under the paths are the links representing the topology. Path names are left.
		\textbf{(a)} \textit{odgi viz} default modality: The colored bars are the paths versus pangenome sequence in a binary matrix. Shown is the subgraph extracted with \textit{odgi extract}. No 1D layout algorithm was applied here.
		\textbf{b-d} \textit{odgi viz} colored by path position. Light grey corresponds to the beginning of a path, black encodes the end of the path.
		\textbf{(b)} The nodes are arranged randomly in 1D.
		\textbf{(c)} The nodes are arranged applying the 1D PG-SGD algorithm.
		\textbf{(d)} The nodes are arranged applying a 1D reference-guided PG-SGD where the nodes of haplotype HG01071 are fixed and only all the othere ones are movable in 1D. Now all paths of this haplotype are arranged from their lowest to their highest nucleotide position. However,lot's of longer links are now visible compared to the node ordering directly above. This indicates a node order that is globally not optimal.}
	\label{fig:1d_sorts}
\end{figure*}

>>>>

\begin{figure*}[!htpb]
	\centering
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.png}
		\label{fig:1d_fig1}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.png}
		\label{fig:1d_fig2}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.Y.og.png}
		\label{fig:1d_fig3}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.HY.og.png}
		\label{fig:1d_fig4}
	\end{subfigure}	
	\caption{\textit{odgi viz} 1D visualizations of a 5 haplotypes subgraph of the Human Pangenome Reference Consortium (HPRC) chromosome 6 pangenome graph. The major histocompatibility complex (MHC) sequence was injected as an extra path. Various node arrangements are shown. 
		\textbf{a-d} A graphs nodes are arranged from left to right forming the pangenome sequence. The black lines under the paths are the links representing the topology. 
		Path names are left.
		\textbf{(a)} \textit{odgi viz} default modality: The colored bars are the paths versus pangenome sequence in a binary matrix. 
		Shown is the subgraph extracted with \textit{odgi extract}. No 1D layout algorithm was applied here.
		\textbf{b-d} \textit{odgi viz} colored by path position. 
		Light grey corresponds to the beginning of a path, black encodes the end of the path.
		\textbf{(b)} The nodes are arranged randomly in 1D.
		\textbf{(c)} The nodes are arranged applying the 1D PG-SGD algorithm.
		\textbf{(d)} The nodes are arranged applying a 1D reference-guided PG-SGD where the nodes of haplotype HG01071 are fixed and only all the othere ones are movable in 1D. 
		Now all paths of this haplotype are arranged from their lowest to their highest nucleotide position. 
		However,lot's of longer links are now visible compared to the node ordering directly above. This indicates a node order that is globally not optimal.}
	\label{fig:1d_sorts}
\end{figure*}

### 2023-09-22 manuscript/fig_2d_layouts.tex
add 1D figure, cleaning
integrate feedback from Niklas

\begin{figure*}[!htb]
	\centering
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/hprc-v1.0-pggb.og.lay.H4000.r4.0-highlighted.png}
		\label{fig:sfig1}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.MHC_annotated.png}
		\label{fig:sfig2}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.MHC_genes_annotated.png}
		\label{fig:sfig3}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.C4_gene_annotated.png}
		\label{fig:sfig4}
	\end{subfigure}	
	\caption{2D visualizations of all chromosomes of the Human Pangenome Resource Consortium (HPRC) 90 haplotypes pangenome graph, chromosome 6, the major histocompatibility complex (MHC), and the complement component 4 (C4). 
		\textbf{(a)} \textit{odgi draw} layout of the HPRC pangenome graph 90 haplotypes. 
		Displayed are all 24 autosomes and the mitochondrial chromosome.
		A red rectangle highlights chromosome 6 which is shown in the subfigure below. 
		\textbf{(b)} \textit{gfaestus} screenshot of the chromosome 6 layout. 
		Colored in blue is the MHC. 
		The hairball in the middle is the centromere. 
		The black structures in the centromere are edges. 
		\textbf{(c)} \textit{gfaestus} screenshot of the MHC. 
		All MHC genes are color annotated and the names of the genes appear as a text overlay. 
		\textbf{(d)} \textit{gfaestus} screenshot of the region around C4, specifically color highlighting genes C4A and C4B. 
		The black lines are the edges of the graph.}
	\label{fig:2d_layouts}
\end{figure*}

>>>>

\begin{figure*}[!htpb]
	\centering
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/hprc-v1.0-pggb.og.lay.H4000.r4.0-highlighted.png}
		\label{fig:2d_fig1}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.MHC_annotated.png}
		\label{fig:2d_fig2}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.MHC_genes_annotated.png}
		\label{fig:2d_fig3}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.C4_gene_annotated.png}
		\label{fig:2d_fig4}
	\end{subfigure}	
	\caption{2D visualizations of all chromosomes of the Human Pangenome Reference Consortium (HPRC) 90 haplotypes pangenome graph, chromosome 6, the major histocompatibility complex (MHC), and the complement component 4 (C4). 
		\textbf{(a)} \textit{odgi draw} layout of the HPRC pangenome graph 90 haplotypes. 
		Displayed are all 24 autosomes and the mitochondrial chromosome.
		A red rectangle highlights chromosome 6 which is shown in the subfigure below. 
		\textbf{(b)} \textit{gfaestus} screenshot of the chromosome 6 layout. 
		Colored in blue is the MHC. 
		The hairball in the middle is the centromere. 
		The black structures in the centromere are edges. 
		\textbf{(c)} \textit{gfaestus} screenshot of the MHC. 
		All MHC genes are color annotated and the names of the genes appear as a text overlay. 
		\textbf{(d)} \textit{gfaestus} screenshot of the region around C4, specifically color highlighting genes C4A and C4B. 
		The black lines are the edges of the graph.}
	\label{fig:2d_layouts}
\end{figure*}

### 2023-09-22 manuscript/main.tex
integrate feedback from Niklas

	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to move pairs of nodes in parallel with a modified HOGWILD! strategy.
	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
	To our knowledge, no algorithm takes into account such biological information to compute the graph layouts.
	PG-SGD can be extended in any number of dimensions.
	In the ODGI toolkit \citep{Guarracino2022}, we provide implementations for 1-dimensional (1D) and 2-dimensional (2D) layouts. The respective algorithms were already successfully applied to build and visualize large scale pangenome graphs of the Human Pangenome Resource Consortium (HPRC)~\citep{Liao2023, Guarracino2023}.

@@

	We report PG-SGD's pseudocode in Algorithm~\ref{alg:pg_sgd} and its schematic in Figure~\ref{fig:sketches}.
	In brief, the algorithm moves one pair of nodes $( v_i, v_j )$ at a time, minimizing the difference between the layout distance $ld_{ij}$ of the two nodes and the nucleotide distance $nd_{ij}$ of the same nodes as calculated along a path that traverses them.
	In the 2D layouts, nodes have two ends.
	When moving a pair of nodes, we actually move one end of each node.
	$v_i$ is the node associated with the step $s_i$ sampled uniformly from all the steps in $\mathcal{P}$.
	$v_j$ is the node associated with the step $s_j$ sampled from the same path of $s_i$ by drawing a Zipfian distribution \citep{Zipf1932}.
	The difference between $nd_{ij}$ and $ld_{ij}$ guides the update of the node coordinates in the layout.
	The magnitude $r$ of the update depends on the learning rate $\mu$.
	The number of iterations steers the annealing step size $\eta$ which determines the learning rate $\mu$.
	%A key property of pangenome graph layouts is that their nodes can be arranged so that most edges go between nodes that are close together in the node order \citep{Guarracino2022}.
	A large $\eta$ in the first iterations leads to a globally linear (in 1D) or planar (in 2D) layout.
	By decreasing $\eta$, the layout adjustments become more localized, ensuring that the nodes are positioned to best reflect the nucleotide distances in the paths (i.e., in the genomes).

@@

    \subsection{Pangenome graphs layouts reveal biology features}
	Graph visualization is essential for understanding pangenome graphs and the genome variation they represent. 
	We show how 2D PG-SGD allows us gaining insight into biological data by looking at the graph layout structure. 
	%To demonstrate this, we provide various 2D PG-SGD layouts in Figure~\ref{fig:2d_layouts}.
			\input{fig_2d_layouts}
	In Figure~\ref{fig:2d_fig1}, the chromosomes of the HPRC graph show the large scale structural variations in the centromeres.
	Focusing on the major histocompatibility complex (MHC) of chromosome 6 (Figure~\ref{fig:2d_fig2}), the 2D layout reveals the positions and diversity of all MHC genes (Figure~\ref{fig:2d_fig3}).
	In Figure~\ref{fig:2d_fig4} the C4A and C4B genes are highlighted. 
	%The layout is the same as the one produced by \textit{Bandage}~\citep{Wick_2015} and displayed in~\citep{Guarracino2022}.
	Complementary, we provide various 1D visualizations of in Supplementary Figure~\ref{fig:1d_sorts}.

@@

	Graph visualization is key for understanding genome variations and the layouts produced by PG-SGD offer an unprecedented high-level perspective on pangenome variation.
	We implemented PG-SGD to generate layouts in 1D and 2D.
	These graph projections have already been employed in constructing and analyzing the first draft human pangenome reference \citep{Liao2023}, as well as in the discovery of heterologous recombination of human acrocentric chromosomes \citep{Guarracino2023}.
	Furthermore, they are applied in the creation and analysis of pangenome graphs for any species \citep{Guarracino2022, Garrison2023}.
	%The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
	%A prerequisite is that the graph nodes are sorted according to their occurrence in the graph's embedded paths.
	%Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
	Of note, there still remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
	Our algorithm will underpin new pangenome graph browsers for studying graph layouts and the genome variation they represent (\url{https://github.com/chfi/waragraph)} last accessed Jul 2023).

>>>>

	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to move pairs of nodes in parallel with a modified HOGWILD! strategy.
	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
	To our knowledge, no algorithm takes into account such biological information to compute the graph layouts.
	PG-SGD can be extended in any number of dimensions.
	In the ODGI toolkit \citep{Guarracino2022}, we provide implementations for 1-dimensional (1D) and 2-dimensional (2D) layouts. The respective algorithms were already successfully applied to build and visualize large scale pangenome graphs of the Human Pangenome Reference Consortium (HPRC)~\citep{Liao2023, Guarracino2023}.

@@

	We report PG-SGD's pseudocode in Algorithm~\ref{alg:pg_sgd} and its schematic in Figure~\ref{fig:sketches}.
	In brief, the algorithm moves one pair of nodes $( v_i, v_j )$ at a time, minimizing the difference between the layout distance $ld_{ij}$ of the two nodes and the nucleotide distance $nd_{ij}$ of the same nodes as calculated along a path that traverses them.
	In the 2D layouts, nodes have two ends.
	When moving a pair of nodes, we actually move one end of each node. For clarification, an example is given in Figure~\ref{fig:sketches}.
	$v_i$ is the node associated with the step $s_i$ sampled uniformly from all the steps in $\mathcal{P}$.
	$v_j$ is the node associated with the step $s_j$ sampled from the same path of $s_i$ by drawing a uniform or a Zipfian distribution \citep{Zipf1932}.
	The difference between $nd_{ij}$ and $ld_{ij}$ guides the update of the node coordinates in the layout.
	The magnitude $r$ of the update depends on the learning rate $\mu$.
	The number of iterations steers the annealing step size $\eta$ which determines the learning rate $\mu$.
	%A key property of pangenome graph layouts is that their nodes can be arranged so that most edges go between nodes that are close together in the node order \citep{Guarracino2022}.
	A large $\eta$ in the first iterations leads to a globally linear (in 1D) or planar (in 2D) layout.
	By decreasing $\eta$, the layout adjustments become more localized, ensuring that the nodes are positioned to best reflect the nucleotide distances in the paths (i.e., in the genomes).

@@

    \subsection{Pangenome graph layouts reveal biology features}
	Graph visualization is essential for understanding pangenome graphs and the genome variation they represent. 
	We show how 2D PG-SGD allows us gaining insight into biological data by looking at the graph layout structure. 
	%To demonstrate this, we provide various 2D PG-SGD layouts in Figure~\ref{fig:2d_layouts}.
			\input{fig_2d_layouts}
	In Figure~\ref{fig:2d_fig1}, the chromosomes of the HPRC graph show the large scale structural variations in the centromeres.
	Focusing on the major histocompatibility complex (MHC) of chromosome 6 (Figure~\ref{fig:2d_fig2}), the 2D layout reveals the positions and diversity of all MHC genes (Figure~\ref{fig:2d_fig3}).
	In Figure~\ref{fig:2d_fig4} the C4A and C4B genes are highlighted. 
	%The layout is the same as the one produced by \textit{Bandage}~\citep{Wick_2015} and displayed in~\citep{Guarracino2022}.
	Complementary, we provide various 1D visualizations in Supplementary Figure~\ref{fig:1d_sorts}.

@@

	Graph visualization is key for understanding genome variations and the layouts produced by PG-SGD offer an unprecedented high-level perspective on pangenome variation.
	We implemented PG-SGD to generate layouts in 1D and 2D.
	These graph projections have already been employed in constructing and analyzing the first draft human pangenome reference \citep{Liao2023}, as well as in the discovery of heterologous recombination of human acrocentric chromosomes \citep{Guarracino2023}.
	Furthermore, they are applied in the creation and analysis of pangenome graphs for any species \citep{Guarracino2022, Garrison2023}.
	%The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
	%A prerequisite is that the graph nodes are sorted according to their occurrence in the graph's embedded paths.
	%Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
	Of note, there still remains a gap in interactive and scalable solutions that merge layouts of large pangenome graphs with annotation.
	Our algorithm will underpin new pangenome graph browsers for studying graph layouts and the genome variation they represent (\url{https://github.com/chfi/waragraph)} last accessed Jul 2023).
